A Simple Type Theory for Category Theory

### 2022-04-23 paper/paper.tex
overview of the algebraic semantics

\begin{definition}[Category with Families]
  A category with families consists of
  \begin{enumerate}
  \item A category $\textrm{Ctx}\isaCat$
  \item A presheaf $\textrm{Ty} : \harr {\textrm{Ctx}}{1}$
  \item A presheaf $\textrm{Tm} : \harr {\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}} \Gamma *} 1$
  \item A terminal object in $\textrm{Ctx}$
  \item An \emph{extension operation}
    $\textrm{ext} : \Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}\Gamma * \to \textrm{Ctx}$
    right representing the profunctor
    \[ \Delta:\textrm{Ctx};A:\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}{\Gamma}{*} \vdash \tensorexistsXwithYandZ {a}{(\punitinXfromYtoZ {} \Delta {\pi^-\pi^-a})}
       {(\punitinXfromYtoZ {}{\pi^-a}{A})}\]
       % {\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}{\Gamma}{*}}
    % 
  \end{enumerate}
\end{definition}

@@

To simplify our initiality proof, we work with an \emph{essentially
algebraic} presentation of the notion of VE hyperdoctrine. As with all
GATs, this has an initial model, which is a kind of ``combinator''
syntax with explicit substitution for VE hyperdoctrines. Then we show
that our type theoretic syntax is equivalent to the combinator syntax
by a simple inductive ``simplification'' algorithm. We conjecture that
a full normalization proof is possible, which would establish useful
conservativity results for the calculus.

The theory has sorts Ctx, Subst, Tm, Ty, Cat, Fun, Prof, ProfCtx and
Trans.

>>>>

Since many type theories are given semantics by free structured
categories, we can use SCT as a logical framework in which to do
``formal type theory''.

\begin{definition}[Theory of unary Computational lambda calculus with booleans]
  \begin{enumerate}
  \item A category $C$
  \item A terminal object $1$ of $C$
  \item A monad $T$ on $C$
  \item A type $\textrm{Bit} : \varr 1 C$
  \item Morphisms $\textrm{y},\textrm{n} : \pendallXdotY * \punitinXfromYtoZ {1(*)} {\textrm{Bit}(*)}$
  \item A natural transformation $\textrm{if} : \pendallXdotY * \homrallXYtoZ \alpha {(\punitinXfromYtoZ C {1(*)} \alpha) \times (\punitinXfromYtoZ C {1(*)} \alpha)} {\punitinXfromYtoZ C {\textrm{Bit}(*)} \alpha}$
  \item Satisfying $\textrm{if} y = \pi_1$ and $\textrm{if} n = \pi_2$
  \item A morphism $\textrm{print} : \pendallXdotY * \punitinXfromYtoZ {\textrm{Bit}(*)}{T (1(*))}$
  \end{enumerate}
\end{definition}

Then we can ``program'' in this computational lambda calculus within
our logical framework by constructing terms of type $\punitinXfromYtoZ
C {A} {B}$ for any functors $A,B$ into $C$.

Using a terminal category and the category of elements, we can define
a formal category with families, an algebraic formulation of the
judgmental structure of dependent type theory:
\begin{definition}[Category with Families]
  A category with families consists of
  \begin{enumerate}
  \item A category $\textrm{Ctx}\isaCat$
  \item A presheaf $\textrm{Ty} : \harr {\textrm{Ctx}}{1}$
  \item A presheaf $\textrm{Tm} : \harr {\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}} \Gamma *} 1$
  \item A terminal object in $\textrm{Ctx}$
  \item An \emph{extension operation}
    $\textrm{ext} : \Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}\Gamma * \to \textrm{Ctx}$
    right representing the profunctor
    \[ \Delta:\textrm{Ctx};A:\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}{\Gamma}{*} \vdash \tensorexistsXwithYandZ {a}{(\punitinXfromYtoZ {} \Delta {\pi^-\pi^-a})}
       {(\punitinXfromYtoZ {}{\pi^-a}{A})}\]
       % {\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}{\Gamma}{*}}
    % 
  \end{enumerate}
\end{definition}

@@

To simplify our initiality proof, we use an \emph{essentially
algebraic} presentation of our notion of model. To describe this, we
give a ``combinator'' syntax for VE hyperdoctrines, where we have
explicit sorts for contexts $\Gamma$ and $\Phi$ and explicit judgments
for substitutions $\gamma : \Delta \to \Gamma$ and $\phi : \Psi
\vdash_{f,g} \Phi$.
%
While the explicit description of the theory is quite long, we can
describe it in simple modular pieces.
%
First, we use the standard theory of a category with families, which
describes the algebraic structure of contexts $\Gamma$, types $\Gamma
\vdash A$ and terms $\Gamma \vdash M : A$.
%
Second, we add to that a theory for a slight generalization of virtual
double categories we call \emph{double categories with display
pro-arrows}.
%
The difference is that in a double category with display pro-arrows,
we have a true double category with strict vertical and horizontal
composition, among whom are the ``display'' pro-arrows.
%
The display pro-arrows model the profunctors of the syntax, which
don't have a strict composition, whereas the arbitrary pro-arrows
model the contexts $\Phi$, which do have a strict composition, given
by context concatenation $\Phi,\Psi$.
%
The difference is that in a virtual double category the contexts are
assumed to be given inductively by formal compositions of the display
pro-arrows, whereas in models of this algebraic theory, the contexts
might include other exotic objects.
%
This difference is analogous to the difference between a category with
families and a contextual category in models of dependent type theory,
and we fully expect that there is a similar relationship between the
models: virtual double categories should form a co-reflective
subcategory characterized as those double categories with display
pro-arrows where all pro-arrows are equal to some concatenation of
displays.
%
Finally, we parameterize all of the constructions of the theory of
double categories with pro-arrows by contexts $\Gamma$ and add in
equations that substitution commutes with all double category
operations.
%
This means the theory presents a category with families with an double
category with display pro-arrows internal to the category of
presheaves on the category of contexts, i.e., a hyperdoctrine of
double categories with display pro-arrows.

The main advantage of this algebraic approach is that the initiality
theorem for the combinatory syntax is trivial. Instead, the work is to
show that our type theoretic syntax, which restricts combinatory rules
to make substitution an admissible rather than primitive operation, is
equivalent to the combinatory syntax.

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma : \isaCtx}

    \inferrule
    {\Delta : \isaCtx \and \Gamma : \isaCtx}
    {\gamma : \textrm{Subst}(\Delta,\Gamma)}

    \inferrule
    {\Gamma : \isaCtx}
    {C : \textrm{Cat}(\Gamma)}

    \inferrule
    {\Gamma : \isaCtx \and C : \textrm{Cat}(\Gamma) \and D : \textrm{Cat}(\Gamma) \and
    }
    {F : \textrm{Functor}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \isaCtx \and C : \textrm{Cat}(\Gamma) \and D : \textrm{Cat}(\Gamma) \and
    }
    {P : \textrm{Profunctor}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \isaCtx \and C : \textrm{Cat}(\Gamma) \and D : \textrm{Cat}(\Gamma)
    }
    {P : \textrm{ProfCtx}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \textrm{Cat}(\Gamma) \and D  : \textrm{Cat}(\Gamma) \and
      C' : \textrm{Cat}(\Gamma) \and D' : \textrm{Cat}(\Gamma) \and
      f : \textrm{Functor}(\Gamma,C,C') \and g : \textrm{Cat}(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      \Psi : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {\phi : \textrm{ProfSubst}(\Gamma,C,D,C',D',f,g,\Phi,\Psi)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \textrm{Cat}(\Gamma) \and D  : \textrm{Cat}(\Gamma) \and
      C' : \textrm{Cat}(\Gamma) \and D' : \textrm{Cat}(\Gamma) \and
      f : \textrm{Functor}(\Gamma,C,C') \and g : \textrm{Cat}(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      P : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {t : \textrm{Trans}(\Gamma,C,D,C',D',f,g,\Phi,P)}
  \end{mathpar}
  \caption{Sorts of Algebraic SCT}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \cdots
  \end{mathpar}
  \caption{Algebraic Dependent Type Theory}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule
    {g : \textrm{Functor}(D,E) \and \textrm{Functor}(C,D)}
    {g \circ_{\Gamma,C,D,E} f : \textrm{Functor}(C,E)}

    (h \circ g) \circ f = h \circ (g \circ f)

    \inferrule
    {}
    {\id_C : \textrm{Functor}(C,C)}

    (h \circ \id) = h

    \id \circ h = h

    \inferrule
    {P : \textrm{Profunctor}(\Gamma,C',D')
    \and f : \textrm{Functor}(\Gamma,C,C')\and
    \and g : \textrm{Functor}(\Gamma,D,D')}
    {P \circ_{\Gamma,C,D,C',D'} (f,g) : \textrm{Profunctor}(\Gamma,C,D)}

    P \circ (f,g) \circ (f',g') = P \circ (f \circ f', g \circ g')

    P \circ (\id,\id) = P

    \inferrule
    {\psi : \textrm{ProfSubst}(f',g',\Phi,\Psi)\and \phi : \textrm{ProfSubst}(f,g,\Sigma,\Phi)
    }
    {\psi \circ \phi : \textrm{ProfSubst}(f' \circ f, g' \circ g, \Sigma, \Psi)}

    \inferrule
    {}
    {restrict(f,g,\phi) : \textrm{ProfSubst}(\id,\id,\Psi,\Phi)}
  \end{mathpar}
  \caption{Algebraic Presentation of SCT}
\end{figure}
